Can Auxiliary Indicators Improve COVID-19 Forecasts and Hotspot Prediction?

### 2021-06-13 forecast/paper/paper.tex
Edits up thru methods D (formerly C)

% Please add a significance statement to explain the relevance of your work
% TODO write one. This is the indicator's one.
\significancestatement{To study the COVID-19 pandemic, its effects on society,
  and measures for reducing its spread, researchers need detailed data on the
  course of the pandemic. Standard public health data streams suffer
  inconsistent reporting and frequent, unexpected revisions. They also miss
  other aspects of a population's behavior, worthy of consideration. We
  present an open database of COVID signals in the United States, measured at
  the county level and updated daily. These include traditionally reported COVID
  cases and deaths, and many others: signals based on measures of mobility,
  social distancing, internet search trends, self-reported symptoms, and
  patterns COVID-related activity in de-identified medical insurance claims. The
  database provides all signals in a common, easy-to-use format, empowering both
  public health research and operational decision-making.
}

@@

At the outset, we should be clear that our intent in this paper is \textit{not}
to provide an authoritative take on cutting-edge COVID-19 forecasting methods.
Instead, our purpose here is to provide a rigorous, quantitative assessment of
the utility that several auxiliary indicators---such as those derived from
internet surveys or medical insurance claims---provide in tasks that involve
predicting future trends in confirmed COVID-19 cases. To assess such utility in
as simple terms as possible, we center our study in the framework of a basic
autoregressive model (in which COVID cases in the near future are predicted from  
a linear combination of COVID cases in the near past), and ask whether the 
inclusion of an auxiliary indicator as an exogenous feature in such a model
improves predictive power. While forecasting carries a rich literature that
offers a wide range of techniques, see e.g., \cite{Hyndman:2018}, we purposely
constrain ourselves to very simple models, avoiding common enhancements such as
order selection, correction of outliers/anomalies in the data, and inclusion of
regularization or nonlinearities. That said, analyses of forecasts submitted to
the COVID-19 Forecast Hub \cite{ForecastHub} by a large community of modelers
have shown that simple, robust models have consistently been among the
best-performing over the pandemic \cite{Cramer:2021}, including time series
models similar to those we consider in what follows.

@@

For more details on the five indicators (including how these are precisely
computed from the underlying data streams) we refer to
\url{https://cmu-delphi.github.io/delphi-epidata/api/covidcast_signals.html},
which documents all of the COVIDcast signals, as well our companion paper on 
COVIDcast API and database. For CTIS in particular, we refer to our companion 
paper on this survey. For the Google Search Trends data set, see
\cite{GoogleSymptoms}, and \cite{Klopfen:2020, Vaira:2020} for articles on the
relevance of anosmia and ageusia.

As for geographic resolution, we consider the prediction of COVID-19 case rates
and hotspots aggregated at the level of an individual \textit{hospital referral
  region} (HRR). HRRs correspond to groups of counties in the United States
within the same hospital referral system. The Dartmouth Atlas of Healthcare 
Policy~\cite{DartmouthHRR}, defines these 306 regions based on a number of
characteristics. They are contiguous regions containing at least one city where
major procedures (cardiovascular or neurological) are performed.  Most hospital
services for the population in an HRR are performed by hospitals within the
region.  Each HRR has a minimum population of about 120,000; some HRRs are quite
large (such as the one containing Los Angeles, which has about 10 million
people), but generally HRRs are much more homogenous in size than the
(approximately) 3200 counties are, and they serve as a nice middle ground in 
between counties and states.

HRRs, by their definition, would be most relevant for forecasting hospital
admissions.  We have chosen to focus on cases (forecasting and predicting
hotspots) at this resolution since the indicators considered should be more
useful in predicting case versus hospitalization activity, the latter being
intuitively more closely connected (in time) to the events that are measured by
the given five indicators. Predicting case rates (and hotspots) at the HRR level
is still a reasonable goal in its own right; and moreover, it could be used to
feed predicted case information into downstream hospitalization models.

@@

In this paper, all models are fit using only ``vintage'' training data. This
means that for a given prediction date, say September 1, 2020, we train models
using only the data that would have been available to us \textit{as of}
September 1.  Essentially, we ``rewind'' the clock to September 1, and query the
COVIDcast API to get the latest data that it would have had available at
that point in time.  This is made possible by the COVIDcast API's comprehensive
data versioning system (described in more detail in our companion paper).  We
also use the \texttt{evalcast} R package \cite{EvalcastR}, which streamlines the
process of training arbitrary prediction models over a sequence of prediction
dates, by automatically constructing the (properly-versioned) sequence of
vintage training data sets.

Vintage training data means different things, in practice, for different
signals. The three signals based on medical claims, CHNG-CLI, CHNG-COVID, and
DV-CLI, are most often 3-5 days latent, and subject to a considerable degree of
revision or ``backfill'' after their initial publication date.  The survey-based
signal, CTIS-CLI-in-community, is 2 days latent, and rarely undergoes any 
revision at all.  The target itself, reported case rates (or a binary hotspot
variable, defined based on reported case rates), is 1 day latent, and exhibits
frequent, unpredictable revisions (sometimes clear anomalies) after initial
publication.  Compared to the revision pattern in the medical claims signals,
which are appears to be much more systematic, the revision pattern in the case
rates signal can be highly erratic, resulting in big spikes as reporting
backlogs are cleared, changes in case definitions are made, other corrections
are applied, etc.

TODO What is ground truth??
TODO Better example of backfill in DV-CLI?

@@

%\subsubsection{Forecasting Commonalities} % PNAS wants Title Caps for everything

On every day between \attn{July 1 and December 31}, we make two types of
forecasts for each of the 306 hospital referral regions in the United States:
(1) we forecast the seven-day trailing average of incident cases, or "smoothed
incident cases"; and (2) we predict whether the HRR will be a hotspot, a
location with a large increase in cases. For both tasks we try to predict the
outcome 7, 8, 9, up to \attn{21} days into the future. For both tasks, we
estimate a baseline global model across all HRRs using the most recent
observation of the target, as well the observations from 1 and 2 weeks earlier.
We train a separate model for each forecast date and for each forecast horizon,
but we do not assume any HRR-specific effects; that is, a single global model is
used to 
describe all HRRs for each forecast date-horizon pair. This model is
re-estimated for every forecast date and each future target resulting in
\attn{$180\times 15 = 2700$} estimated baseline models for each task.  
We then add current and lagged values (analogous to the baseline) for each of
the four signals separately. We now provide further details on each of these
tasks.

\subsubsection{Task 1: Case Forecasting}

To fix notation, let $Y_{\ell,t}$ denote the smoothed COVID-19 case
incidence rate per 100,000 people for location (HRR) $\ell$ and time (day) $t$. 
Let $X_{\ell,t}$ denote
the one of the indicators described above, for example the Facebook \%
CLI-in-community signal, 
for location $\ell$ and time $t$.
We evaluate five models of the form:
\begin{align*}
Y_{\ell,t+d}
&\approx \alpha + \sum_{j=0}^2 \beta_j Y_{\ell,t-7j} \\
Y_{\ell,t+d}
&\approx \alpha + \sum_{j=0}^2 \beta_j Y_{\ell,t-7j} +
\sum_{j=0}^2 \gamma_j X_{\ell,t-7j}, 
\end{align*}
where we also exchange $X_{\ell, t}$ for the signals from \attn{CHNG and
  Google.} 
Here $d=7,\ldots, 21$ depending on the target value
(number of days ahead we predict).

Informally, the first model, which we will call the ``Cases'' model, 
bases its predictions of future case rates on the following three features:
current COVID-19 case rates and those 1 and 2 weeks back.
The second model, ``Cases + Facebook'', additionally incorporates the 
current Facebook signal and the Facebook signal from 1 and 2 weeks in the past.
The third model, ``Cases + Google'', is exactly the same but substitutes the 
Google signal instead of the Facebook one and similarly for ``Cases + Change''.

For each model, in order to make a forecast at time $t_0$
(to predict case rates at time $t_0+d$),
we estimate the unknown parameters using quantile
regression~\cite{KoenkerXiao2006}, 
training over all locations $\ell$,
and all times $t$ that are within the most recent 21 days of data
available up to and including time $t_0$.
Rather than predicting only the median (or mean) value for each forecast date
and each horizon, we estimate seven quantiles, coinciding 
with the county-level forecast submission guidelines for the
COVID-19 Forecast Hub~\cite{Cramer:2021}.  These are \{0.025, 0.1,
0.25, 0.5, 
0.75, 0.9, 0.975\}. This provides a notion of uncertainty in the forecasts. An
example of this type of forecast is shown in \autoref{fig:trajectory}.

\begin{figure}[tb!]
    \centering
    \includegraphics[width=.9\linewidth]{fig/trajectory.pdf}
    \caption{Forecasts and observed cases for HRR x. \attn{Create a figure here
        for 1 HRR. Show the fan for the quantiles.}} 
    \label{fig:trajectory}
\end{figure}

% Noncrossing is enforced via post hoc ordering.

@@

We also examine whether alternative indicators help in predicting hotspots. On a
given date, we say an HRR is a hotspot if the average number of observed cases
over the past 7 days has increased by at least $25$\% compared to the preceding
week. We remove all data where, on average, fewer than $30$ cases were observed
over the preceding $7$ days. Thus, hotspot status can be represented by a
$\{0,1\}$-valued variable, and hotspot prediction is a binary classification
problem.

As in the case forecasting task, in hotspot prediction we train models that use
previously observed values of cases and alternative indicators as features to
predict a response (in our case, hotspot status). In addition to the differently
defined response, there are two other differences between the hotspot prediction
and case forecasting tasks. First, in hotspot prediction we always use relative
changes in previous observed cases/indicators as features. That is, rather than
$Y_{t,\ell}$ we use 
$$
Y^\Delta_{\ell,t} = \frac{Y_{\ell,t} - Y_{\ell,
    t-7}}{Y_{\ell,t-7}}\quad\text{and}\quad X^\Delta_{\ell,t} =
\frac{X_{\ell,t} - X_{\ell, t-7}}{X_{\ell,t-7}} 
$$ 
Second, because hotspot prediction is a binary classification problem, we
estimate the parameters $\alpha$, $\beta$ and $\gamma$ by logistic regression
rather than quantile regression.

@@

Code to reproduce all examples (which uses the
\texttt{evalcast} R package and fetches data directly from the API) can be
found at \url{TODO}.

>>>>

% Please add a significance statement to explain the relevance of your work
% TODO write a real one. This is the indicator's one.
\significancestatement{To study the COVID-19 pandemic, its effects on society,
  and measures for reducing its spread, researchers need detailed data on the
  course of the pandemic. Standard public health data streams suffer
  inconsistent reporting and frequent, unexpected revisions. They also miss
  other aspects of a population's behavior, worthy of consideration. We
  present an open database of COVID signals in the United States, measured at
  the county level and updated daily. These include traditionally reported COVID
  cases and deaths, and many others: signals based on measures of mobility,
  social distancing, internet search trends, self-reported symptoms, and
  patterns COVID-related activity in de-identified medical insurance claims. The
  database provides all signals in a common, easy-to-use format, empowering both
  public health research and operational decision-making.
}

@@

At the outset, we should be clear that our intent in this paper is \textit{not}
to provide an authoritative take on cutting-edge COVID-19 forecasting methods.
Instead, our purpose here is to provide a rigorous, quantitative assessment of
the utility that several auxiliary indicators---such as those derived from
internet surveys or medical insurance claims---provide in tasks that involve
predicting future trends in confirmed COVID-19 cases. To assess such utility in
as simple terms as possible, we center our study in the framework of a basic
autoregressive model (in which COVID cases in the near future are predicted from  
a linear combination of COVID cases in the near past), and ask whether the 
inclusion of an auxiliary indicator as an additional feature in such a model
improves predictive power. While forecasting carries a rich literature that
offers a wide range of techniques, see e.g., \cite{Hyndman:2018}, we purposely
constrain ourselves to very simple models, avoiding common enhancements such as
order selection, correction of outliers/anomalies in the data, and inclusion of
regularization or nonlinearities. That said, analyses of forecasts submitted to
the COVID-19 Forecast Hub \cite{ForecastHub} by a large community of modelers
have shown that simple, robust models have consistently been among the
best-performing over the pandemic \cite{Cramer:2021}, including time series
models similar to those we consider in what follows.

@@

For more details on the five indicators (including how these are precisely
computed from the underlying data streams) we refer to
\url{https://cmu-delphi.github.io/delphi-epidata/api/covidcast_signals.html},
which documents all of the signals in the COVIDcast API, and our companion paper
on the API and database. For CTIS in particular, we refer to our companion paper
on this survey. For the Google Search Trends data set, see
\cite{GoogleSymptoms}, and \cite{Klopfen:2020, Vaira:2020} for articles on the
relevance of anosmia or ageusia (loss of smell or taste) to COVID.

As for geographic resolution, we consider the prediction of COVID-19 case rates
and hotspots aggregated at the level of an individual \textit{hospital referral
  region} (HRR). HRRs correspond to groups of counties in the United States
within the same hospital referral system. The Dartmouth Atlas of Healthcare 
Policy~\cite{DartmouthHRR}, defines these 306 regions based on a number of
characteristics. They are contiguous regions such that most of the hospital
services for the underlying population are performed by hospitals within the
region. Each HRR also contains at least one city where major procedures
(cardiovascular or neurological) are performed. The smallest HRR has a
population of about 120,000. While some are quite large (such as the one
containing Los Angeles, which has about 10 million people), generally HRRs 
are much more homogenous in size than the (approximately) 3200 counties,  
and they serve as a nice middle ground in between counties and states.

HRRs, by their definition, would be most relevant for forecasting hospital
demand.  We have chosen to focus on cases (forecasting and predicting
hotspots) at the HRR level since the indicators considered should be more 
useful in predicting case activity versus hospital demand, since the latter is
intuitively more closely connected (in time) to the events that are measured by
the given five indicators. Predicting case rates (and hotspots) at the HRR level
is still a reasonable goal in its own right; and moreover, it could be used to
feed predicted case information into downstream hospitalization models.

@@

In this paper, all models are fit with ``vintage'' training data. This means
that for a given prediction date, say, September 1, 2020, we train models 
using only the data that would have been available to us \textit{as of}
September 1 (imagine that we ``rewind'' the clock to September 1, and query the
COVIDcast API to get the latest data it would have had available at that point
in time.)  This is made possible by the COVIDcast API's comprehensive data  
versioning system (described in more detail in our companion paper).  We also
use the \texttt{evalcast} R package \cite{EvalcastR}, which streamlines the
process of training arbitrary prediction models over a sequence of prediction
dates, by constructing the proper sequence of vintage training data sets.

Vintage training data means different things, in practice, for different
signals. The three signals based on medical claims, CHNG-CLI, CHNG-COVID, and
DV-CLI, are most often 3-5 days latent, and subject to a considerable but
regular degree of revision or ``backfill'' after their initial publication date.
The survey-based signal, CTIS-CLI-in-community, is 2 days latent, and rarely
undergoes any revision at all.  The target variable itself, reported COVID-19
case rates, is 1 day latent, and exhibits frequent, unpredictable revisions
(sometimes clear anomalies) after initial publication.  Compared to the revision
pattern in the medical claims signals, which are appears to be much more
systematic in nature, the revision pattern in the case rates signal can be
highly erratic, resulting in big spikes as reporting backlogs are cleared,
changes in case definitions are made, other corrections are applied, etc.

TODO Are the above problems mentioned about case rates really to do with vintage 
data, or would they still exist in finalized data?

TODO Should we discuss the use of ``bad'' case rates as ground truth??

TODO Discuss Google-AA.  We're using all finalized data, but explain why we're
OK with this ... (they never backfill).

TODO For a figure, use Jacob's ``famous'' example of backfill in DV-CLI?   
\url{https://cmu-delphi.github.io/covidcast/talks/ml-summit/talk.html#(20)}

\subsection{Missing Data Imputation}

TODO Briefly describe that we zero-fill Google-AA.

@@

The first task we consider---\textit{forecasting}---is to predict the HRR-level
COVID-19 case incidence rates, averaged over some 7-day time period in the
future (the average number of new daily reported cases per 100,000 people, over
the 7-day period).  The second task---\textit{hotspot prediction}---is to
predict a binary variable defined in terms of relative changes of the
aforementioned target variable. Why define the target variable here via 7-day 
averaging? The short answer is robustness: averaging stabilizes the case time
series, and accounts for uninteresting artifacts like weekend-weekday
differences in the series.  Note that we can of course equivalently view this
(equivalent up to a constant factor) as predicting HRR-level COVID-19 case
incidence rates \textit{summed} over some 7-day period in the future (the total
number of new reported cases per 100,000 people, over the 7-day period), or
predicting a binary variable defined in terms of its relative changes.

To fix notation, let $Y_{\ell,t}$ denote the 7-day trailing average of COVID-19
case incidence rates in location (HRR) $\ell$ and at time (day) $t$.  In 
forecasting, a time $t$, we seek to predict $Y_{\ell,t+a}$ for each ``ahead''
value $a=7,\ldots,21$. Meanwhile, in hotspot prediction, we seek to predict
$Z_{\ell,t+a}$, a binary indicator defined in terms of the relative difference 
$(Y_{\ell,t+a}-Y_{\ell,t+a-7}) / Y_{\ell,t+a-7}$, again for each
$a=7,\ldots,21$.

For each prediction date $t$, we use a 21-day trailing window of data to train
our forecast or hotspot prediction models (so, e.g., the trained models will
differ from those at prediction date $t-1$).  This is done to account for
(potential) nonstationarity.  For simplicity, the forecasting and hotspot
prediction models are always fit to data across all HRRs (i.e., the coefficients
in the models do not account for location-specific effects).

In our analysis, we let the prediction date $t$ run over each day in between
early June and December 31, 2020.  The precise start date differs for
forecasting and hotspots prediction; for each task it was chosen to be the
earliest date at which the data needed to train all models was available, which
ends up being (per our setup, with 21 days of training data, and lagged values
of signals for features as we will detail shortly) June 6, 2020 for forecasting,
and June 13, 2020 for hotspot prediction. (The bottleneck here is the
CTIS-CLI-in-community signal, which does not exist before early April 2020, when
the survey was first launched). The end date was chosen with a consideration
again to align both tasks as best as possible, and because few hotspots exist
post December 31, 2020, due to the general and gradual decline of the pandemic
in 2021.

\subsubsection{Task 1: Forecasting}

Recall $Y_{t,\ell}$ denotes the 7-day trailing average of COVID-19 case
incidence rates in location $\ell$ and at time $t$.  Separately for each
$a=7,\ldots,28$, to predict $Y_{\ell,t+a}$ for ahead value $a$, we consider a
simple probabilistic forecasting model of the form:     
\begin{equation}
\label{eq:forecast_ar}
\Quantile_\tau(Y_{\ell,t+a} \,|\, Y_{\ell,s}, s \leq t)  
= \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y_{\ell,t-7j}.  
\end{equation}
This model uses current case rates, and the case rates 7 and 14 days ago, in
order to predict (the quantiles of) case rates in the future.  We consider a
total of 7 quantile levels (chosen in accordance with the county-level quantile
levels suggested by the COVID-19 Forecast Hub),      
\begin{equation}
\label{eq:quantile_levels}
\tau \in \{0.025, 0.1, 0.25, 0.5, 0.75, 0.9, 0.975 \}.
\end{equation}
To estimate the coefficients, we use \textit{quantile regression}
\cite{Koenker:1978, Koenker:2005, Koenker:2006} fit separately for each $\tau$,
using data from all 306 HRRs, and within each HRR, using the most recent 21 days
of training data.  To be clear, this gives 6,426 training samples for each
quantile regression problem.

In addition to the above ``pure'' autoregression model, we consider five
probabilistic forecasting models of the form: 
\begin{multline}
\label{eq:forecast_ar_x}
\Quantile_\tau(Y_{\ell,t+a} \,|\, Y_{\ell,s}, X_{\ell,s}, s \leq t)  
= \\ \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y_{\ell,t-7j} + 
\sum_{j=0}^2 \gamma^{a,\tau}_j X_{\ell,t-7j},
\end{multline}
where $X_{\ell,t}$ denotes any one of the five auxiliary indicators---CHNG-CLI, 
CHNG-COVID, CTIS-CLI-in-community, DV-CLI, or Google-AA---at location $\ell$ and
time $t$. Note that we apply the same lags (current value, along with the values
7 and 14 days ago) for the auxiliary signals as we do for case rates. Estimation
proceeds just as before: we use the same 7 quantile levels as in
\eqref{eq:quantile_levels}, and again use quantile regression separately for
each $\tau$, fit using data from all 306 HRRs and a trailing window of 21 days
of training data.

At prediction time, in order to deal with possible crossing violations (a
predicted quantile at level $\tau$ exceeds a predicted quantile at level $\tau'
> \tau$), we apply a simple post-hoc sorting.  See Figure \ref{fig:trajectory}
for example forecast.

\begin{figure}[tb!]
\centering
\includegraphics[width=\columnwidth]{fig/trajectory.pdf}
\caption{Forecasts and observed cases for HRR x. TODO 
  \attn{Create a figure here for 1 HRR. Show the fan for the quantiles.}}  
\label{fig:trajectory}
\end{figure}

@@

Define the binary indicator:
$$
Z_{t,\ell} = \indicator{Y^\Delta_{\ell,t} \geq 0.25},
$$
where we use the notation \smash{$Y^\Delta_{\ell,t} = (Y_{\ell,t} - Y_{\ell,
    t-7})/(Y_{\ell,t-7})$}. In other words, $Z_{t,\ell}=1$ if the number of
newly reported cases over the past 7 days has increased by at least 25\%
compared to the preceding week.  When this occurs, we call location $\ell$ a
\textit{hotspot} at time $t$.

We treat hotspot prediction as a binary classification problem and use a setup 
altogether quite similar to the forecasting setup described previously.
Separately for each $a=7,\ldots,28$, to predict $Z_{\ell,t+a}$, we consider a
simple logistic model:
\begin{equation}
\label{eq:hotspot_ar}
\logit(Z_{\ell,t+a} \,|\, Y_{\ell,s}, s \leq t)  
= \alpha^{a,\tau} + \sum_{j=0}^2 \^{a,\tau}_j Y^\Delta_{\ell,t-7j}.  
\end{equation}
In addition to this ``pure'' autoregression model, we also consider five
logistic models of the form: 
\begin{multline}
\label{eq:hotspot_ar_x}
\logit(Z_{\ell,t+a} \,|\, Y_{\ell,s}, X_{\ell,s}, s \leq t)  
= \\ \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y_{\ell,t-7j} + 
\sum_{j=0}^2 \gamma^{a,\tau}_j X_{\ell,t-7j},
\end{multline}
where we use \smash{$X^\Delta_{\ell,t} = (X_{\ell,t} - X_{\ell,
    t-7})/(X_{\ell,t-7})$}, and again $X_{\ell,t}$ stands for any of the five  
auxiliary indicators at location $\ell$ and time $t$.  We fit all models
\eqref{eq:hotspot_ar}, \eqref{eq:hotspot_ar_x} using logistic regression, 
pooling all 306 HRRs and using a 21-day trailing window for the training data.

An important detail is that in hotspot prediction we remove all data from
training and evaluation where, on average, fewer than 30 cases (this refers to a
count, not a rate) are observed over the preceding 7 days. This avoids having to
make arbitrary calls for a hotspot (or lack thereof) based on small counts.

@@

Code to reproduce all results (which uses the \texttt{evalcast} R package) can 
be found at
\url{https://www.github.com/cmu-delphi/covidcast-pnas/forecast/code/}.
